---
title: "The Impact of Transmission Type on Fuel Economy in the mtcars Dataset"
author: "Justin Z"
date: "April 29, 2019"
output: pdf_document
---


```{r setup, include = FALSE}

knitr::opts_chunk$set(echo = TRUE)

# Create a function to fit, summarize, and sort models from input variables
FitAndSortModels <- function(data, response, predictors.unique,
                             predictors.repeat = NULL) {
  # Assembles formulas to be used as input for lm, fits models, displays summary
  # statistics side-by-side, and sorts the result by decreasing adj. R-squared
  #
  # Args:
  #   data: the data set containing the variables, passed to lm
  #   response: A chr vector of the variable to be predicted by the others
  #   predictors.unique: A chr vector of variables to iterate (fit 1 model each)
  #   predictors.repeat: A chr vector of variables to repeat in all models
  #
  # Returns:
  #   A tibble containing the model formulas, model objects, and summary stats
  #
  # First build a vector of formulas from the character vectors of variables
  formula.vector <- MakeLmInput(response = response,
                                predictors.unique = predictors.unique,
                                predictors.repeat = predictors.repeat)
  # Next make a tibble which stores the model results alongside the formulas
  table.1 <- tibble(model.name = formula.vector) %>%
    mutate(model.object = map(model.name, lm, data = data))
  # Next make a tibble which collects summary statistics for each model
  table.2 <- table.1$model.object %>%  # List of the models
    map(GetModelStats) %>%  # Apply function to each element, output as list
    bind_rows()  # Condense the list down to a tibble
  # Lastly bind the two tibbles together and sort by decreasing adj. R-Squared
  result <- bind_cols(table.1, table.2) %>%
    arrange(desc(Ad.R.Squared))
  result  # Return the tibble
}

# Create a function to assemble lm formulas from input varibles
MakeLmInput <- function(response, predictors.unique, predictors.repeat = NULL) {
  # Assembles character vectors into formulas to be used as input for lm
  #
  # Args:
  #   response: A chr vector of the variable to be predicted by the others
  #   predictors.unique: A chr vector of variables to iterate (fit 1 model each)
  #   predictors.repeat: A chr vector of variables to appear in all models
  #
  # Returns:
  #   A character vector of "formulas" which can be coerced into actual
  #   formulas by lm.
  #
  # The pasting is slightly different depending if any variables will be
  # repeated in all the models, the if statement below handles this
  if (is.null(predictors.repeat)) {
    repeated.portion <- paste(response, "~") # Looks like "mpg ~"
    separator <- " "  # Additional varible is added to the right with a space
  } else {
    repeated.portion <- paste(predictors.repeat, collapse = " + ") %>%
      paste(response, ., sep = " ~ ")  # Looks like "mpg ~ var"
    separator <- " + "  # Additonal variable is added to the right with "+"
  }
  # The variables are pasted below, the repeated portion is recycled for each of
  # the unique predictors. The output vector has length(predictors.unique)
  paste(repeated.portion, predictors.unique, sep = separator)
}

# Create a function to collect summary statistics for an individual model object
GetModelStats <- function(model) {
  # Returns a tibble of stats for given model object
  #
  # Args:
  #   model: an object containing results returned by a model function (e.g. lm)
  #
  # Returns:
  #   A tibble object of the R-Squared, Adj. R-Squared, and p-value of the model
  tibble(R.Squared = summary(model)$r.squared,
         Ad.R.Squared = summary(model)$adj.r.squared,
         model.pvalue = anova(model)$`Pr(>F)`[[1]])  # Easier to get from anova
}

```


## Overview

This report is for the peer-graded project in the Regression Models course from
Johns Hopkins University within the Data Science Specialization on Coursera. The
instructions say to perform exploratory data analysis and use regression models
with the `mtcars` data set to answer these questions:

>1. Is an automatic or manual transmission better for fuel economy
2. Quantify the impact of transmission type on fuel economy


## Executive Summary

The objective of this analysis was to determine the impact of transmission type
on fuel economy in the `mtcars` data set. The data were loaded into R and
pre-processed. Exploratory data analysis was performed, and potential
interaction effects were identified. Model selection progressed in stages, and
the best model for predicting fuel economy was found to take the form of
`mpg ~ hp + wt + hp * wt`. Adding the `am` variable to this model did not
significantly improve the model, so it was concluded that transmission type does
not have a significant impact on fuel economy in this data set. While the impact
was quantified, it is important to note that it is likely due to random chance.


## Part 1) Loading and Pre-processing the Data

Before beginning it should be noted that the assignment instructions specify a
maximum length for this report, so in the interest of conserving space some text
and figure output will be suppressed throughout the report. When output is
suppressed it will be noted at the beginning of the code chunk, and the output
will be described as comments. Additionally, three functions were written to
faciliate model selection, and to conserve space the code for these functions is
not shown in this report. If interested the functions can be viewed in the
GitHub repo for this project submission here.

The first step in the analysis loads the necessary packages and data.

```{r loading, message = FALSE}

# Messages suppressed for this code chunk
library(datasets)  # For the mtcars data
library(tidyverse)
library(car)  # For the VIF function

# Load and view some information about the data
data("mtcars")
mtcars <- as_tibble(mtcars, rownames = "vehicle")  # Put row names in a column
str(mtcars)  # Check structure, all numeric besides vehicle names

```

Except for the vehicle names, all the variables came through as numeric, but it
is clear that there is a mix of numeric, categorical, and discrete variables
here. The help document can provide additional information on the descriptions,
units, and coding of the factor variables. The next code chunk will assess the
discrete variables and convert the data types as needed.

```{r preprocessing, fig.show = "hide"}

# Figure output is suppressed in this code chunk
# Check if the discrete variables should be treated as numeric or categorical
par(mfrow = c(3, 2))  # Setup plot space
boxplot(mtcars$mpg ~ mtcars$cyl)
plot(mtcars$mpg ~ mtcars$cyl)  # Only three levels, but decreases mostly linear
boxplot(mtcars$mpg ~ mtcars$gear)
plot(mtcars$mpg ~ mtcars$gear)  # Only three levels, shaped like inverted V
boxplot(mtcars$mpg ~ mtcars$carb)
plot(mtcars$mpg ~ mtcars$carb)  # Six levels, and decreasing somewhat linearly
# Based on the checks above, cyl and carb will be treated as numeric and gear
# will be treated as categorical

# Proceed with converting data types
mtcars <- mtcars %>%
  mutate(vs = as.factor(vs)) %>%
  mutate(am = as.factor(am)) %>%
  mutate(gear = as.factor(gear))

# The vs and am variables are coded with zeros and ones, but it will be easier
# to work with them by the names those values represent:
levels(mtcars$vs) <- c("V-shaped", "Straight")
levels(mtcars$am) <- c("Automatic", "Manual")

```


## Part 2) Exploratory Data Analysis

Next the variables will be checked for NA values, variation, and covariation.

```{r EDA.variation, results = "hide", fig.show = "hide"}

# In this code chunk, text and figure output is suppressed
# Check for NA values and check variation of the variables
summary(mtcars)  # No NA values; also disp, hp, and carb have means > median
par(mfrow = c(1, 2))  # Setup plot space
hist(mtcars$mpg)  # Nothing unusual, fewer cars in 25 bin, could be noise
plot(as.factor(mtcars$cyl))  # Only three possible values: 4, 6, and 8
hist(mtcars$disp)  # Right skew/flat with 4 obs per bin of 50 wide
hist(mtcars$hp)  # Right skew, fewer observations of large hp
hist(mtcars$drat)  # Looks normally distributed
hist(mtcars$wt)  # Only one car with weight between 4k and 5k, and then 3 at 5k
hist(mtcars$qsec)  # Looks normally distributed
# Engine shape is fairly evenly split: 18 & 14 observations each
# Transmission type is faily evenly split: 19 & 13 observations each
# Gear is mostly 3s and 4s, with only 5 observations in the 5 gear category
plot(as.factor(mtcars$carb))  # Nearly all 1s, 2s, and 4s, then 1 each at 6 & 8

```

The previous chunk provided general information about each varible on its own,
and in the next section covariation of the variables will be checked. First
consider the original question of how trans type relates to mpg.

```{r EDA.cov1}

par(mfrow = c(1, 2))  # Setup plot space
boxplot(mtcars$mpg ~ mtcars$am, ylab = "mpg [mpg]")  # mpg vs trans type
boxplot(mtcars$wt ~ mtcars$am, ylab = "weight [klbs]")  # wt vs. trans type

```

In the first plot above one could jump to the conclusion that `mpg` is clearly
higher with manual transmission. However in the second plot one can see that
manual transmission vehicles tend to be lighter in this dataset, and common
sense suggests that `mpg` should be higher for lower weight vehicles. Many of
these variables can be expected to correlate with `mpg` and/or with each other.
The regression model must correct for these interactions and isolate the effect
of the transmission type on fuel economy.

The `vif` function can be used to demonstrate how correlated the variables are
to eachother. In the code chunk below, one can see that the variance inflation
factors range from 4.98 to 23.5, so these variables tend to be highly correlated
with each other.

```{r EDA.cov2, fig.show = "hide"}

# In this code chunk figure output is suppressed
# Check the variance inflation factors for all variables 
vifs <- lm(mpg ~ cyl + disp + hp + drat + wt + qsec + vs + am + gear + carb,
           data = mtcars) %>%  # Fit a model of all variables
  vif()  # Apply vif function to the model
vifs[, 1] %>%
  summary() %>%
  print()
rm(vifs)

# Check how mpg varies with the other variables
# It was shown above that mpg decreases with number of cylinders, as expected
plot(mtcars$mpg ~ mtcars$disp)  # Negative slope, to be expected
plot(mtcars$mpg ~ mtcars$hp)  # Negative slope, to be expected
plot(mtcars$mpg ~ mtcars$drat)  # Positively sloped, opposite of expected
plot(mtcars$mpg ~ mtcars$wt)  # Negative slope, very much expected
plot(mtcars$mpg ~ mtcars$qsec)  # Generally positive, faster cars consume more
plot(mtcars$mpg ~ mtcars$vs)  # Straight appears to have a significantly higher
# mpg, but such a strong trend is not expected
# It was shown above that mpg appears higher for manual cars, but this could be
# misleading based on the other factors
# It was shown above that there doesnt appear to be a strong trend with mpg vs
# number of gears, but one isn't really expected either
# It was shown above that there is a negative slope between mpg and carb, but
# such a strong relationship isn't really expected

```

A few of the interesting notes above were followed up in the analysis. In the
interest of conserving space those details will not be outlined here. If
interested the complete EDA portion can be found on the GitHub repo for this
project submission here. In summary, the unexpected results above are no longer
present once other factors are controlled for. In the final EDA code chunk a few
additional covariations are explored.

```{r EDA.cov3, fig.show = "hide"}

# In this code chunk figure output is suppressed
# Exploratory plots of interactions
plot(mtcars$disp ~ mtcars$am)  # Lower for manual, but expected considering wt
plot(mtcars$hp ~ mtcars$disp)  # Positive slope, to be expected
plot(mtcars$hp ~ mtcars$wt)  # Positive slope, to be expected
plot(mtcars$hp ~ mtcars$am)  # Lower for manual but expected if considering wt
plot(mtcars$drat ~ mtcars$am)  # Higher for manual, but again the wt factor
plot(mtcars$qsec ~ mtcars$am)  # Minimal difference, to be expected

```

It is also worth mentioning these two points:

* By definition, displacement = effective cylinder volume * number of cyls
* Quarter mile time is a result of many features of the car similar to mpg,
while it can be expected to correlate with mpg, it is likely to be very
correlated with all the other variables. This variable should be treated with
caution when considering models.

Based on the checks above and general knowledge about the variables, seventeen
interaction effects will be considered in the analysis. The complete list can be
found in the model selection section below.


## Part 3a) Model Selection, Round 1

The model was built up sequentially with the idea of forward selection in mind.
First the variables were checked to see which had the strongest relationship
with `mpg`. To facilitate model selection a function called `FitAndSortModels`
was written which takes variables as input, fits models, collects summary
statistics, and sorts the models by performance. The code for this function is
not shown, but its input and output is shown below.

```{r MS.round1}

# First check which variable has the strongest relationship with mpg by itself
# Collect all the predictor variables into a character vector
round1.vars <- mtcars %>%
  select(-vehicle, -mpg) %>%  # Drop the non-predictor variables
  names()
# Fit, summarize, and sort a series of models
round1.table <- FitAndSortModels(data = mtcars, response = "mpg",
                                 predictors.unique = round1.vars)
print(round1.table)  # mpg ~ wt is the top model
round1.model <- round1.table$model.object[[1]]  # Store the model object

# Check out the residuals of the mpg ~ wt model
par(mfrow = c(1, 2))  # Setup plot space
hist(residuals(round1.model))  # Right skew, ranges from -4.54 to 6.87
plot(residuals(round1.model) ~ mtcars$wt)  # V shape high at ends low in center
rm(round1.vars, round1.table, round1.model)

```

Due to the shape of the residuals plot above, log transforms of the wt variable
were explored. To save space those details are omitted here, but they can be
found in the GitHub repo. While the log transforms did improve the model very
slightly, ultimately the simpler model was selected.


## Part 3b) Model Selection, Round 2

The next round of model selection will consider interaction effects along with
the main variables.

```{r MS.round2a}

# Build a character vector of the interaction terms
round2.interactions <- c("cyl * disp", "disp * hp", "disp * wt", "disp * qsec",
                          "hp * wt", "hp * carb", "hp * qsec", "drat * wt",
                          "drat * qsec", "wt * qsec", "wt * vs", "wt * am",
                          "wt * gear", "wt * carb", "qsec * am", "qsec * gear",
                          "qsec * carb")
# Consolidate into a single character vector of variables
round2.vars <- mtcars %>%
  select(-vehicle, -mpg, -wt) %>%  # Reduce to only the repeated vars
  names() %>%  # Extract main variables
  c(round2.interactions)  # Add in interaction effects
# Fit, summarize, and sort a series of models
round2.table <- FitAndSortModels(data = mtcars, response = "mpg",
                                 predictors.unique = round2.vars,
                                 predictors.repeat = "wt")
print(round2.table)  # mpg ~ wt + hp + wt * hp is the top model
round2.model <- round2.table$model.object[[1]]  # Store the model object
rm(round2.interactions, round2.vars, round2.table)

```

The code chunk above shows that `mpg ~ wt + hp + hp * wt` is the top performing
model from this round. Next the `anova` function will be used to see if the
addition of the extra terms is significant. Then the residuals will be checked
as well.

```{r MS.round2b}

# Check anova results for the nested model series: mpg ~ wt + hp + hp * wt
test1 <- lm(mpg ~ wt, mtcars)
test2 <- lm(mpg ~ wt + hp, mtcars)
test3 <- lm(mpg ~ wt + hp + hp * wt, mtcars)
print(anova(test1, test2, test3))  # The test3 model is still significant
rm(test1, test2, test3)
# Anova suggests keeping all of the terms

# Check out the residuals
par(mfrow = c(1, 3))  # Setup plot space
hist(residuals(round2.model))  # Right skew, ranges from -3.06 to 4.55
plot(residuals(round2.model) ~ mtcars$wt)  # Looks like random noise
plot(residuals(round2.model) ~ mtcars$hp)  # Somewhat triangular

```

The `anova` results above suggest keeping all three terms in the model. No
significant pattern is noticed in the residuals either. An additional round of
model selection was performed where 24 models were fit each with one additional
term to the round 2 model. None of the additions improved the model enough to
pass an Anova test, but to conserve space these results will not be shown here.
Now that the model has been identified the original questions can be addressed.


## Part 4) Conclusions

The assignment instructions ask:

>1. Is an automatic or manual transmission better for fuel economy
2. Quantify the impact of transmission type on fuel economy

This is somewhat of a trick question since the data suggest that the type of
transmission does not have a significant impact on fuel economy. This can be
shown by adding `am` to the model and checking `anova`.

```{r conclusion}

part4.model <- lm(mpg ~ hp + wt + hp * wt + am, data = mtcars)
print(summary(part4.model))  # coef of amManual is 0.13, pvalue is 0.9259
print(anova(round2.model, part4.model))  # p-value 0.93

```

The code chunk above shows that once the `wt` and `hp` variables are controlled
for that the impact of `am` is both small and insignificant, and with a pvalue
of 0.9259 there is little uncertainty about this. If one includes the variable
anyways there appears to be a 0.13 mpg improvement when using a manual
transimission instead of automatic, but this observation is very likely due to
chance.

