---
title: "The Impact of Transmission Type on Fuel Economy in the mtcars Dataset"
author: "Justin Z"
date: "April 29, 2019"
output: pdf_document
---


```{r setup, include = FALSE}

knitr::opts_chunk$set(echo = TRUE)

# Create a function to fit, summarize, and sort models from input variables
FitAndSortModels <- function(data, response, predictors.unique,
                             predictors.repeat = NULL) {
  # Assembles formulas to be used as input for lm, fits models, displays summary
  # statistics side-by-side, and sorts the result by decreasing adj. R-squared
  #
  # Args:
  #   data: the data set containing the variables, passed to lm
  #   response: A chr vector of the variable to be predicted by the others
  #   predictors.unique: A chr vector of variables to iterate (fit 1 model each)
  #   predictors.repeat: A chr vector of variables to repeat in all models
  #
  # Returns:
  #   A tibble containing the model formulas, model objects, and summary stats
  #
  # First build a vector of formulas from the character vectors of variables
  formula.vector <- MakeLmInput(response = response,
                                predictors.unique = predictors.unique,
                                predictors.repeat = predictors.repeat)
  # Next make a tibble which stores the model results alongside the formulas
  table.1 <- tibble(model.name = formula.vector) %>%
    mutate(model.object = map(model.name, lm, data = data))
  # Next make a tibble which collects summary statistics for each model
  table.2 <- table.1$model.object %>%  # List of the models
    map(GetModelStats) %>%  # Apply function to each element, output as list
    bind_rows()  # Condense the list down to a tibble
  # Lastly bind the two tibbles together and sort by decreasing adj. R-Squared
  result <- bind_cols(table.1, table.2) %>%
    arrange(desc(Ad.R.Squared))
  result  # Return the tibble
}

# Create a function to assemble lm formulas from input varibles
MakeLmInput <- function(response, predictors.unique, predictors.repeat = NULL) {
  # Assembles character vectors into formulas to be used as input for lm
  #
  # Args:
  #   response: A chr vector of the variable to be predicted by the others
  #   predictors.unique: A chr vector of variables to iterate (fit 1 model each)
  #   predictors.repeat: A chr vector of variables to appear in all models
  #
  # Returns:
  #   A character vector of "formulas" which can be coerced into actual
  #   formulas by lm.
  #
  # The pasting is slightly different depending if any variables will be
  # repeated in all the models, the if statement below handles this
  if (is.null(predictors.repeat)) {
    repeated.portion <- paste(response, "~") # Looks like "mpg ~"
    separator <- " "  # Additional varible is added to the right with a space
  } else {
    repeated.portion <- paste(predictors.repeat, collapse = " + ") %>%
      paste(response, ., sep = " ~ ")  # Looks like "mpg ~ var"
    separator <- " + "  # Additonal variable is added to the right with "+"
  }
  # The variables are pasted below, the repeated portion is recycled for each of
  # the unique predictors. The output vector has length(predictors.unique)
  paste(repeated.portion, predictors.unique, sep = separator)
}

# Create a function to collect summary statistics for an individual model object
GetModelStats <- function(model) {
  # Returns a tibble of stats for given model object
  #
  # Args:
  #   model: an object containing results returned by a model function (e.g. lm)
  #
  # Returns:
  #   A tibble object of the R-Squared, Adj. R-Squared, and p-value of the model
  tibble(R.Squared = summary(model)$r.squared,
         Ad.R.Squared = summary(model)$adj.r.squared,
         model.pvalue = anova(model)$`Pr(>F)`[[1]])  # Easier to get from anova
}

```


## Overview

This report is for the peer-graded project in the Regression Models course from
Johns Hopkins University within the Data Science Specialization on Coursera. The
instructions say to perform exploratory data analysis and use regression models
with the `mtcars` data set to answer these questions:

>1. Is an automatic or manual transmission better for fuel economy
2. Quantify the impact of transmission type on fuel economy


## Executive Summary

The objective was to determine the impact of transmission type on fuel economy
in the `mtcars` data set. The data were loaded into R, pre-processed, and then
exploratory data analysis was performed. The model selection phase aimed to
identify the best predictive model for fuel economy so that the role of
transmission type could be assessed. Model selection progressed in stages, and
the best model was found to take the form `mpg ~ hp + wt + hp * wt`. Adding the
`am` variable to this model did not significantly improve the model, so it was
concluded that transmission type does not have a significant impact on fuel
economy in this data set.


## Part 1) Loading and Pre-processing the Data

In order to meet the report length requirement some text and figure output will
be suppressed, when applicable this is noted at the beginning of the code chunk,
and the output will be described as comments. Additionally, 3 functions were
written to facilitate model selection, the code for these is not shown either.
If interested, unabriged analysis can be viewed in the [GitHub repo][1] for this
project submission.

[1]: https://github.com/jtzingsheim1/mtcars-Fuel-Economy "GitHub repo"

The first step in the analysis loads the necessary packages and data.

```{r loading, message = FALSE}

# Messages suppressed for this code chunk
library(datasets); library(tidyverse); library(car)

data("mtcars")  # Load the data
mtcars <- as_tibble(mtcars, rownames = "vehicle")  # Put row names in a column
str(mtcars)  # Check structure, all numeric besides vehicle names, no NA values

```

Except for the vehicle names, all the variables came through as numeric, but it
is clear that there is a mix of numeric, categorical, and discrete variables
here. The help document can provide additional information on the descriptions,
units, and coding of the factor variables. After assessing the varibles it was
decided to treat `cyl` and `carb` as numeric and `gear` as categorical, so
conversions occur accordingly below.

```{r preprocessing}

# Proceed with converting data types
mtcars <- mtcars %>%
  mutate(vs = as.factor(vs)) %>%
  mutate(am = as.factor(am)) %>%
  mutate(gear = as.factor(gear))

# Convert the zeros and ones of cateogrical variables
levels(mtcars$vs) <- c("V-shaped", "Straight")
levels(mtcars$am) <- c("Automatic", "Manual")

```


## Part 2) Exploratory Data Analysis

Next the variation of the variables was checked, but to save space that code is
not shown here, if interested the complete EDA can be found in the
[GitHub repo][1]. A key portion of this project is dealing with the covariation
of the variables as shown in the next code chunk.

```{r EDA.cov1}

par(mfrow = c(1, 2))  # Setup plot space
boxplot(mtcars$mpg ~ mtcars$am, ylab = "mpg [mpg]")  # mpg vs trans type
boxplot(mtcars$wt ~ mtcars$am, ylab = "weight [klbs]")  # wt vs. trans type

```

In the first plot above one could jump to the conclusion that `mpg` is clearly
higher with manual transmission. However in the second plot one can see that
manual transmission vehicles tend to be lighter in this dataset, and common
sense suggests that `mpg` should be higher for lower weight vehicles. Many of
these variables can be expected to correlate with `mpg` and/or with each other.
The regression model must correct for these interactions and isolate the effect
of the transmission type on fuel economy. The variance inflation factors for
these variables were checked, and the values were quite high, confirming that
these variables are all very much related. In the next code chunk the
covariations with mpg were checked.

```{r EDA.cov2}

# These plots are commented out to conserve space
# Check how mpg varies with the other variables
#plot(mtcars$mpg ~ mtcars$cyl)  # Negative slope, as expected
#plot(mtcars$mpg ~ mtcars$disp)  # Negative slope, to be expected
#plot(mtcars$mpg ~ mtcars$hp)  # Negative slope, to be expected
#plot(mtcars$mpg ~ mtcars$drat)  # Positively sloped, opposite of expected
#plot(mtcars$mpg ~ mtcars$wt)  # Negative slope, very much expected
#plot(mtcars$mpg ~ mtcars$qsec)  # Generally positive, faster cars consume more
#plot(mtcars$mpg ~ mtcars$vs)  # Strong relationship, but that is not expected
#plot(mtcars$mpg ~ mtcars$am)  # Strong relationship, but that is not expected
#plot(mtcars$mpg ~ mtcars$gear)  # Weak relationship, as expected
#plot(mtcars$mpg ~ mtcars$carb)  # Negative relationship, but not expected

```

A few of the interesting notes above were followed up in the analysis, and the
unexpected trends disappear when accounting for the significant variables.
Additional covariations were explored, but that code is not shown here. Based on
the checks above and general knowledge about the variables, 17 interaction
effects will be considered in the analysis (complete list below).


## Part 3a) Model Selection, Round 1

The model was built up sequentially with the idea of forward selection in mind.
First the variables were checked to see which had the strongest relationship
with `mpg`. To facilitate model selection a function called `FitAndSortModels`
was written which takes variables as input, fits models, collects summary
statistics, and sorts the models by performance. The code for this function is
not shown, but its input and output is shown below.

```{r MS.round1}

# First check which variable has the strongest relationship with mpg by itself
# Collect all the predictor variables into a character vector
round1.vars <- mtcars %>%
  select(-vehicle, -mpg) %>%  # Drop the non-predictor variables
  names()
# Fit, summarize, and sort a series of models
round1.table <- FitAndSortModels(data = mtcars, response = "mpg",
                                 predictors.unique = round1.vars)
print(head(round1.table, n = 4))  # mpg ~ wt is the top model
round1.model <- round1.table$model.object[[1]]  # Store the model object

# Check out the residuals of the mpg ~ wt model
par(mfrow = c(1, 2))  # Setup plot space
hist(residuals(round1.model))  # Right skew, ranges from -4.54 to 6.87
plot(residuals(round1.model) ~ mtcars$wt)  # V shape high at ends low in center
rm(round1.vars, round1.table, round1.model)

```

Due to the shape of the residuals plot above, log transforms of the wt variable
were explored. To save space those details are omitted here, but they can be
found in the [GitHub repo][1]. While the log transforms did improve the model
very slightly, ultimately the simpler model was selected.


## Part 3b) Model Selection, Round 2

The next round of model selection will consider interaction effects along with
the main variables.

```{r MS.round2a}

# Build a character vector of the interaction terms
round2.interactions <- c("cyl * disp", "disp * hp", "disp * wt", "disp * qsec",
                          "hp * wt", "hp * carb", "hp * qsec", "drat * wt",
                          "drat * qsec", "wt * qsec", "wt * vs", "wt * am",
                          "wt * gear", "wt * carb", "qsec * am", "qsec * gear",
                          "qsec * carb")
# Consolidate into a single character vector of variables
round2.vars <- mtcars %>%
  select(-vehicle, -mpg, -wt) %>%  # Reduce to only the repeated vars
  names() %>%  # Extract main variables
  c(round2.interactions)  # Add in interaction effects
# Fit, summarize, and sort a series of models
round2.table <- FitAndSortModels(data = mtcars, response = "mpg",
                                 predictors.unique = round2.vars,
                                 predictors.repeat = "wt")
print(head(round2.table, n = 4))  # mpg ~ wt + hp + wt * hp is the top model
round2.model <- round2.table$model.object[[1]]  # Store the model object
rm(round2.interactions, round2.vars, round2.table)

```

The code chunk above shows that `mpg ~ wt + hp + hp * wt` is the top performing
model from this round. The `anova` function and the residuals were checked for
this model.

```{r MS.round2b, results = "hide"}

# Text results are hidden for this code chunk
# Check anova results for the nested model series: mpg ~ wt + hp + hp * wt
test1 <- lm(mpg ~ wt, mtcars)
test2 <- lm(mpg ~ wt + hp, mtcars)
test3 <- lm(mpg ~ wt + hp + hp * wt, mtcars)
anova(test1, test2, test3)  # Anova suggests keeping all of the terms
rm(test1, test2, test3)

# Check out the residuals
par(mfrow = c(1, 3))  # Setup plot space
hist(residuals(round2.model))  # Right skew, ranges from -3.06 to 4.55
plot(residuals(round2.model) ~ mtcars$wt)  # Looks like random noise
plot(residuals(round2.model) ~ mtcars$hp)  # Somewhat triangular

```

The `anova` results above suggest keeping all three terms in the model. No
significant pattern is noticed in the residuals either. An additional round of
model selection was performed where 24 models were fit each with one additional
term to the round 2 model. None of the additions improved the model enough to
pass an Anova test, but to conserve space these results will not be shown here.
Now that the model has been identified the original questions can be addressed.


## Part 4) Conclusions

The assignment instructions ask:

>1. Is an automatic or manual transmission better for fuel economy
2. Quantify the impact of transmission type on fuel economy

This is somewhat of a trick question since the data suggest that the type of
transmission does not have a significant impact on fuel economy. This can be
shown by adding `am` to the model and checking `anova`.

```{r conclusion}

part4.model <- lm(mpg ~ hp + wt + hp * wt + am, data = mtcars)
print(summary(part4.model))  # coef of amManual is 0.13, pvalue is 0.9259
print(anova(round2.model, part4.model))  # p-value 0.93

```

The code chunk above shows that once the `wt` and `hp` variables are controlled
for that the impact of `am` is both small and insignificant, and with a pvalue
of 0.9259 there is little uncertainty about this. If one includes the variable
anyways there appears to be a 0.13 mpg improvement when using a manual
transimission instead of automatic, but this observation is very likely due to
chance.

